What Determines Stock Prices?

### 2024-04-13 outputs/paper/paper.qmd
update paper and modeling script

---
title: "Effect of Earnings and Dividends on Stock Prices"
subtitle: "Analysis of the Relationship Between Stock Price, Earnings Per Share, and Dividends of the Largest 50 North American Companies from 2013 to 2023"
author: Diana Liu
thanks: "Code and data are available at: https://github.com/Diana-Guanzhi-Liu/Effect-of-Earnings-and-Dividends-on-StockPrices"
date: today
date-format: long
abstract: "This study examines the relationship between Earnings per Share (EPS), dividends, and stock prices in publicly traded companies. Findings indicate a positive correlation between EPS and stock price, with each dollar increase in EPS associated with a $26.59 rise in stock price. Conversely, dividend payments are linked to a decrease in stock price, mitigating the impact of EPS on price. These insights enable investors to predict stock prices, informing their investment strategies, as well as guide managerial decisions balancing pleasing existing shareholders by paying dividends, ensuring long-term growth of the company, and keeping the stock attractive to new investors."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

model_1 <-
  lm(
    Price ~ EPS,
    data = analysis_data
  )

saveRDS(
  model_1,
  file = "~/Stock-Prices/outputs/models/model_1.rds"
)
```

```{r}
#| include: false
#| eval: true
#| warning: false
model_rstanarm_2 <-
  stan_glm(
    formula = Price ~ EPS + Paid_Dividend,
    data = analysis_data,
    family = gaussian(),
    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
    prior_intercept = normal(0, 2.5, autoscale = TRUE),
    prior_aux = exponential(rate = 1, autoscale = TRUE),
    seed = 12345
  )

saveRDS(
  model_rstanarm_2,
  file = "~/Stock-Prices/outputs/models/model_rstanarm_2.rds"
)
```

@@

summary(model_1) 
summary(model_rstanarm_2)
```

@@

modelsummary(
  list(
    "EPS Only" = model_1,
    "With Dividends" = model_rstanarm_2
  ),
  fmt = 2
)
```
$R^2$ and adjusted $R^2$ are terms that explains how well the model is able to explain data, with $R^2$ for single variable models and the more conservative adjusted $R^2$ for multiple variables. Model 1 has a $R^2$ of 0.57, meaning that 57% of the variability observed in stock price is explained by EPS in the regression model. Model 2 can explain 59% of stock price with EPS and dividends. Both indicate that the model can explain stock price reasonably well, but there is around 40% of variability is attributed to other causes. This may also be why the @fig-model-2 does not look perfectly linear and has clustering around 0 to 10 EPS.

@@

prior_summary(model_rstanarm_2)
```

@@

pp_check(model_rstanarm_2) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(model_rstanarm_2) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```
@fig-ppcheckandposteriorvsprior-1 compares actual dependent variable outcomes with distributions from the posterior simulation. We can see that the general shape and distribution are similar but actual data has far higher kurtosis than the simulations.

@@

plot(
  model_rstanarm_2,
  "areas"
)
```

@@

plot(model_rstanarm_2, "trace")

plot(model_rstanarm_2, "rhat")
```
Markov chain Monte Carlo (MCMC) to was used in the model to obtain samples from the posterior distributions of interest. @fig-trace-rhat-1 shows bouncing horizontal lines with overlap, and @fig-trace-rhat-2 shows R-hat close to one. Both indicate that there are no issues with the algorithm.

>>>>

---
title: "What Determines Stock Prices?"
subtitle: "Analysis of the Relationship Between Stock Price, Earnings Per Share, and Dividends of the Largest 50 North American Companies from 2013 to 2023"
author: Diana Liu
thanks: "Code and data are available at: https://github.com/Diana-Guanzhi-Liu/What-Determines-Stock-Prices-"
date: today
date-format: long
abstract: "This study examines the relationship between Earnings per Share (EPS), dividends, and stock prices in publicly traded companies. Findings indicate a positive correlation between EPS and stock price, with each dollar increase in EPS associated with a $26.59 rise in stock price. Conversely, dividend payments are linked to a decrease in stock price, mitigating the impact of EPS on price. These insights enable investors to predict stock prices, informing their investment strategies, as well as guide managerial decisions balancing pleasing existing shareholders by paying dividends, ensuring long-term growth of the company, and keeping the stock attractive to new investors."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

model_1 <- readRDS("~/Stock-Prices/outputs/models/model_1.rds")
model_2 <- readRDS("~/Stock-Prices/outputs/models/model_2.rds")
```

@@

summary(model_1) 
summary(model_2)
```

@@

modelsummary(
  list(
    "EPS Only" = model_1,
    "With Dividends" = model_2
  ),
  fmt = 2
)
```
$R^2$ and adjusted $R^2$ are terms that explains how well the model is able to explain data, with $R^2$ for single variable models and the more conservative adjusted $R^2$ for multiple variables. Model 1 has a $R^2$ of 0.57, meaning that 57% of the variability observed in stock price is explained by EPS in the regression model. Model 2 can explain 59% of stock price with EPS and dividends. Both indicate that the model can explain stock price reasonably well, but there is around 40% of variability is attributed to other causes. This may also be why the @fig-model-2 does not look perfectly linear and has clustering around 0 to 10 EPS.

@@

prior_summary(model_2)
```

@@

pp_check(model_2) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(model_2) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```
@fig-ppcheckandposteriorvsprior-1 compares actual dependent variable outcomes with distributions from the posterior simulation. We can see that the general shape and distribution are similar but actual data has far higher kurtosis than the simulations.

@@

plot(
  model_2,
  "areas"
)
```

@@

plot(model_2, "trace")

plot(model_2, "rhat")
```
Markov chain Monte Carlo (MCMC) to was used in the model to obtain samples from the posterior distributions of interest. @fig-trace-rhat-1 shows bouncing horizontal lines with overlap, and @fig-trace-rhat-2 shows R-hat close to one. Both indicate that there are no issues with the algorithm.
